# Unit 1-8 {-}

```{r settings, echo=FALSE, warning=FALSE, message = FALSE}
knitr::opts_chunk$set(echo=FALSE,  message=FALSE, 
                      warning=FALSE, 
                      fig.width = 5, fig.height = 2,
                      out.width = "100%", fig.align = "center"
                      )
library(tidyverse)
my_theme <- theme_set(theme_minimal()) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(plot.title = element_text(size = 9)) +
  theme(axis.text.x = element_text(size = 9))

```

## Unit 1.8 Measures of Variability from the Median {-}

The focus of this unit is understanding the measure of variability which is based on the median of a data set. A measure of variability is a measure of spread. It seeks to describe how spread out data values are in a data set.

Two measures of variability are the **range** and **IQR**. The range is the maximum value minus the minimum value. The IQR is the difference between the first quartile and third quartile.

**What are quartiles?**

These are the positions in the distribution that partition the data set into quarters. You can think about the quartiles as slicing the data. If you slice a piece of rope 3 times, you cut it into 4 pieces.
Similarly, quartiles cut a list of numbers 3 times, dividing them into 4 pieces that each have an equal numbers of values. 

The three numbers that divide the list into 4 equal parts are called quartiles.

- Q1, the "first quartile"
- Q2, the "second quartile". Q2 is also the median.
- Q3, the "third quartile"
 
### Example {-}

Find the range and IQR of the following data set containing a sample of 14 test scores. The following test scores are from a History class in a high school:

**Test Scores**: 58, 65, 72, 72, 73, 74, 75, 75, 75, 76, 76, 78, 78, 100

```{r mammal_dotplot}

data <- c(58, 65, 72, 72, 73, 74, 75, 75, 75, 76, 76, 78, 78, 100)
df <- as.data.frame(data)
ggplot(df,aes(x=data)) + 
  geom_dotplot(fill="#8eedf7", color="black",dotsize = 0.9) +
  labs(title="Sample of Test Scores",x="Test score", y = "") + my_theme

```

**Solution** 

Here is a dotplot of the values in **Desmos**.  Click on this Desmos link below to see the boxplot and how the median and quartiles can be computed.  The _stats_ command is used to find these statistics.

https://www.desmos.com/calculator/ehhgcrrh49

The **range** of this data set is **100 – 58 = 42**. The range tells us the distance between the lowest and highest values. The downside of the range is that it only uses two data values.
We can see from the dotplot that the majority of the test scores are between 72 and 78.

To find the quartiles by hand, we can order the data from low to high, and then separate it into two halves, a lower half, and an upper half.

<table>
  <tr>
    <th>Lower half</th>
    <th>Upper half</th>
  </tr>
  <tr>
    <td>58, 65, 72, 72, 73, 74, 75</td>
    <td>75, 75, 76, 76, 78, 78, 100</td>
  </tr>
</table>

- The **median** of the data set is **75**, because that is the average of the highest number in the lower half and the lowest number in the upper half. Another name for the median is Q2 (the "second quartile" and). Half (50%) of the values in the sample are less than or equal to Q2=75, and half (50%) of the values in the sample are greater than or equal to Q2=75.

- We can now find the median of each half of the data. These are Q1 and Q2.

- The **first quartile (Q1) ** is the median of the **lower half** of the data. In this example Q1 is **72**. Why 72? Because 25% of the values in the data set are less than or equal to 72, and 75% of the values in the data set are greater than or equal to 72.

- The **third quartile (Q3)** is the median of the **upper half** of the data. In this example Q3 is **76**. Why 76? Because 75% of the values in the data set are less than or equal to 76, and 25% of the values in the data set are greater than or equal to 76.

- The **interquartile range (IQR)** is defined as **Q3 – Q1**.  The IQR is the range of the middle 50% of the data values. The **IQR = 76 – 72 = 4**. So, the middle 50% (middle half) of the test scores are within a range of 4.

### Boxplots {-}

We can summarize the quartiles, minimum value, and maximum value in a list called the **5-number summary**. The 5-number summary of the sample is: 

- Min = 58, Q1 = 72, Q2 = 75, Q3 = 76, Max = 100

A graph of the 5-number summary is called a boxplot.

The boxplot for this sample is:

```{r}
ggplot(df,aes(x=data)) + 
  geom_boxplot(fill="#8eedf7", color="black") + ylim(-.8,.8) +
  labs(title="Sample of Test Scores",x="Test score", y = "") + my_theme

```


The box plot shows everything that we calculated above:

- The range is the entire width of the box plot.

- The IQR (interquartile range) is the width of the box.

- The minimum value is the far-left point (58), and the maximum value is the far-right point (100).

- The left edge of the box is Q1 = 72, the line in the middle of the box is Q2 = 75, and the right edge of the box is Q3 = 76.

### Outliers {-}

The boxplot displays the outliers - the data values that are far from the rest of the grades. The outliers are **58, 65, and 100**.
These values are outliers due to the following rule:

**1.5×IQR Rule for Outliers**

A data value is considered an **outlier** if it is:

- Less than the Q1 – 1.5 × IQR ( _the lower fence_ ) 

- Greater than Q3 + 1.5 × IQR ( _the upper fence_ ) 

In this sample, 

- 1.5 × IQR = 1.5(4) = 6**

- Lower fence = Q1 – 1.5 × IQR = 72 – 6 = **66**  

- Upper fence = Q3 + 1.5 × IQR = 76 + 6 = **82** 

- Any data value that is outside the fences is considered an outlier.  Since the test scores of 58 and 65 are less than the lower fence, and since the test score of 100 is greater than the upper fence, these values are outliers in this sample.

### Stacked Boxplots {-}

Boxplots are very useful for visually comparing the center and variability of two or more data sets.  For example, the dotplot below shows life expectancy of countries based on the continent where the country is located.

```{r stacked_boxplot, fig.width = 5, fig.height = 3.3}
library(gapminder)
gapminder %>% ggplot(aes(x=lifeExp, y=continent, fill=continent)) + 
  geom_boxplot(alpha=0.8) + scale_fill_viridis_d(option = "plasma") +
  theme_minimal() + theme(plot.title = element_text(size = 9)) +
  labs(title="Life Expectancy of Countries by Continent",x="Typical Life Expectancy (in years)", y = "") +
  my_theme
```
We can see that countries in Oceania have a median life expectancy which is greater than the median life expectancies of other continents.  Countries in Africa have the lowest median.  We can also compare variability.  The boxplot for Asia contains the the widest box, so countries in Asia have life expectancies with the greatest interquartile range.

## Summary {-}

Here are some main points from Unit 1.8:

- The range is simple to compute and understand, but it is not a great measure of variability because it is completely influenced by unusual observations (outliers).

- The IQR offers an alternative, relatively simple, measure of variability.

- The IQR is not affected much by outliers and skew. We say that the IQR is resistant to outliers.

- Because the IQR is unaffected by outliers, it allows us to determine whether high or low values are outliers, or not. We use the 1.5×IQR rule and quartiles Q1 and Q3 to determine fences. We call any values that are beyond the fences outliers.

## Practice Problems {-}

1.
2.
3.
